---
title: "A3: Incarceration"
output: html_document
---

### Introduction

I choose the dataset [Smaller dataset about prison/jail population per 100,000 
people (rate) per county/state starting at 1990 in Washington]
(https://github.com/melaniewalsh/Neat-Datasets/raw/main/us-prison-jail-rates-1990-WA.csv)
This dataset provides detailed insights into prison and jail rates in Washington 
State from 1990 to 2018. This dataset is instrumental for analyzing trends over 
nearly three decades, focusing on the rates of incarceration per 100,000 people.
Rates is formatted such that each unit represents 1 individual per 100,000 in 
the population, making the data both accessible and interpretable in terms of 
understanding the scale of incarceration relative to the total population size.
In this report, I analyzed key variables reflecting incarceration rates across 
Washington State, with a specific focus on racial disparities. The variables 
examined include the Total Jail Population Rate, average black jail population 
rate, and other specific rates for Black and White populations, and some other 
relevant variables which help trace trends and evaluate the effectiveness of 
interventions. 


### Summary Information

The summary of jail population rates across Washington State underscores 
significant insights into racial disparities and the dynamics of incarceration. 
In the most recent year of data, 39 counties reported, offering a comprehensive 
overview for analysis. The maximum total jail population rate is about 992.82, 
and the highest black jail population rate is about 31000 in Pend Oreille County. 
The data reveals stark disparities, the maximum jail population rate for Black 
individuals is significantly higher than for White individuals, with the largest 
difference recorded in about 4052.54. Furthermore, the recent average jail 
population rates show that Black individuals are incarcerated at a higher rate 
compared to their White counterparts.The average white jail population rate is 
about 326.48, and the average black jail population rate is about 1052.11. 
The change in the average Black jail population rate over the years, which is 
about -707.576.These findings underscore the need for continued efforts to
address racial inequalities within the criminal justice system of Washington 
State, providing a crucial basis for policymakers and community activists 
focused on reform and equity.

``````{r}
library(dplyr)
prison_wa <- read.csv("https://github.com/melaniewalsh/Neat-Datasets/raw/main/us-prison-jail-rates-1990-WA.csv")

summary_info <- list() 
## Total number of counties in Washington in this dataset
summary_info$num_county <- prison_wa %>%
  filter(year == max(year)) %>%
  summarize(num = n()) %>%
  pull(num)
## The maximum total jail population rate in this dataset
summary_info$max_jail_rate <- prison_wa %>%
  filter(total_jail_pop_rate == max(total_jail_pop_rate, na.rm = TRUE)) %>%
  pull(total_jail_pop_rate)
## The maximum black jail population rate in this dataset
summary_info$max_jail_rate_black <- prison_wa %>%
  filter(black_jail_pop_rate == max(black_jail_pop_rate, na.rm = TRUE)) %>%
  pull(black_jail_pop_rate)
## The county of maximum black jail population rate in this dataset
summary_info$max_jail_rate_black_county <- prison_wa %>%
  filter(black_jail_pop_rate == max(black_jail_pop_rate, na.rm = TRUE)) %>%
  pull(county_name)
## The difference between maximum black jail population rate and maximum white 
## jail population rate
summary_info$diff_max_race_jail_rate <- prison_wa %>%
  filter(year == max(year)) %>%
  summarize(max_diff = max(black_jail_pop_rate, na.rm = TRUE) -
              max(white_jail_pop_rate, na.rm = TRUE)) %>%
  pull(max_diff)
## The average white jail population rate in the recent year in this dataset
summary_info$recent_avg_jail_rate_white <- prison_wa %>%
  filter(year == max(year)) %>%
  summarize(avg_rate = mean(white_jail_pop_rate, na.rm = TRUE)) %>%
  pull(avg_rate)
## The average black jail population rate in the recent year in this dataset
summary_info$recent_avg_jail_rate_black <- prison_wa %>%
  filter(year == max(year)) %>%
  summarize(avg_rate = mean(black_jail_pop_rate, na.rm = TRUE)) %>%
  pull(avg_rate)
## The change of average black jail population rate over years
summary_info$change_avg_black_jail_rate <- prison_wa %>%
  group_by(year) %>%
  summarize(avg_rate = mean(black_jail_pop_rate, na.rm = TRUE)) %>%
  filter(year == max(year) | year == min(year)) %>%
  summarize(change = diff(avg_rate)) %>%
  pull(change)

summary_info
```
### The Dataset

Who collected the data?  

- The data was collected by the Vera Institute of Justice, a nonprofit 
organization that focuses on criminal justice reform.
It was compiled by Census of Jails, Census of State and Federal Adult 
CorrectionalFacilities, FBI Uniform Crime Reporting Program, National 
Corrections Reporting Program, Deaths in Custody Reporting Program, Annual 
Survey of Jails, Centers on Disease Control, U.S. Department of Agriculture, and 
U.S. Census Bureau. Many organizations combined and cooperated together.


How was the data collected or generated?

- The data was generated using several sources, and it was created through the 
collaborative efforts of multiple organizations, which enriched it by 
integrating additional data and information to fill any existing gaps.

Why was the data collected?  

- The data was collected to support the Vera Institute of Justice’s project, 
"In Our Backyards." This project aims to highlight and address the local impacts 
of jail and prison use and to understand how incarceration rates vary 
significantly across different counties and jurisdictions. It seeks to provide 
a more detailed analysis than state-level data, which can obscure local 
variations and issues.

How many observations (rows) are in your data?  

- There is 1131 rows in this data.

How many features (columns) are in the data?  

- There are 23 columns in this data.

What, if any, ethical questions or questions of power do you need to consider when working with this data?  

- When analyzing data on incarceration rates, particularly those that 
distinguish rates by race and demographics, several ethical considerations and 
power dynamics come into play. Privacy and anonymity are crucial, requiring 
rigorous measures to prevent the identification of individuals from the data. 
There's also a significant risk of reinforcing negative stereotypes and biases 
if the data is not handled and interpreted with sensitivity towards the groups 
it represents. The accuracy of the data must be critically assessed to ensure it 
does not reflect or amplify existing systemic biases in law enforcement or data 
collection practices.  Researchers and policymakers must navigate these ethical 
waters with transparency about the data's limitations


What are possible limitations or problems with this data?   (at least 200 words)

- Changes in data collection methods over time or reforms in the justice system 
can affect the consistency of data year over year. For instance, the 
introduction of new reporting requirements or changes in the legal definitions 
of crimes can result in apparent trends that are actually artifacts of these 
changes rather than true changes in behavior or enforcement. The data may not 
perfectly represent the entire population of interest. For example, smaller 
jurisdictions or those with less robust data collection infrastructure might 
be underrepresented. This could skew understandings of regional differences in 
incarceration practices. There is a significant amount of missing data and 
timeframe limitations in the dataset (jail data from 1983 to 2016), which can 
pose a challenge to conducting a thorough and accurate analysis. Since the 
prison data only covers up to 2016, and the dataset contains information up to 
2018, this could lead to a lack of data support when analyzing trends in recent 
years. This discontinuity may make it difficult for researchers to assess trends 
in recent years, such as determining whether there is a downward or upward trend 
in prison population numbers. Large amounts of missing data may bias the results 
of statistical analyses. For example, if data for a particular county are 
missing frequently over a period of years, estimates of the average level or 
growth trend of the prison population in that county may not be accurate. 
Missing data may affect the representativeness of the results of the analysis, 
which may also lead to misinterpretation of particular groups or regions.


### Trends Over Time Chart

The chart visualizes the trends in Black jail population rates in the top 5 
counties in Washington state from 1990 to 2018. These top 5 counties are those 
have the highest average black jail population rate. The chart reveals that 
Black jail population rates have experienced significant volatility over the 
decades in all five counties. This fluctuation could be indicative of various 
external factors influencing incarceration rates, such as changes in law 
enforcement practices, policy reforms, economic conditions, or social movements. 
Most counties show pronounced peaks at certain points, particularly in the late 
1990s and early 2000s, followed by sharp declines. For instance, Columbia County
and Yakima County show sharp peaks around the early 2000s. Post-2010, the trends 
in these counties begin to diverge more noticeably. Some counties like Asotin 
show a resurgence in rates, while others like Lincoln continue to decline. 
Overall, all 5 counties showed a decline in black jail population rate even 
with strong fluctuations.


``````{r}
library(ggplot2)

top_5_county <- prison_wa %>%
  filter(!is.na(black_jail_pop_rate)) %>%
  group_by(county_name) %>%
  summarize(mean_rate = mean(black_jail_pop_rate))%>%
  top_n(5, wt = mean_rate) %>%
  pull(county_name)

data1 <- prison_wa %>%
  group_by(year) %>%
  filter(county_name %in% top_5_county) %>%
  filter(!is.na(black_jail_pop_rate))

ggplot(data = data1, aes(x=year, y=black_jail_pop_rate, group=county_name, 
                         color=county_name)) + 
  geom_line() +
  labs(title = "Black Jail Population Rate Trends In Top Mean 5 Counties",
       x = "Year", y = "Black Jail Population Rate (Unit: 0.01)", 
       color = "County Names") +
  scale_x_continuous(breaks = seq(1990, 2018, by=4))
```



### Variable Comparison Chart

The scatterplot visualizes the relationship between the total jail population 
rate and the Black jail population rate across various counties in Washington 
State, using data from 1990 to the 2018. This chart was chosen to highlight 
how general incarceration trends align or differ specifically with the 
incarceration rates of the Black population, potentially pointing to disparities 
or patterns unique to this demographic group. Pink points represent individual 
data points, where each point indicates a county's combined rates of total and 
Black jail population for a given year. The sky-blue line, derived from a linear 
regression model, helps visualize the overall trend or relationship between the 
two variables. There appears to be a positive correlation between the total jail 
population rate and the Black jail population rate. This suggests that in 
counties or years with higher overall jail populations, the Black jail 
population rate is also higher. Points clustered when both variables are 
relatively low, However, as the total jail population rate increases, the Black 
jail population rate shows significant variability. The spread of the points 
increases with higher total jail population rates, indicating more variability 
in the Black jail population rate in counties with higher overall incarceration 
rates. There are observed several outliers.

``````{r}
data2 <- prison_wa %>%
  filter(!is.na(total_jail_pop_rate)) %>%
  filter(!is.na(black_jail_pop_rate))

ggplot(data = data2, aes(x=total_jail_pop_rate, y=black_jail_pop_rate)) +
  geom_point(color="pink") +
  geom_smooth(method=lm, color="skyblue", se=FALSE) +
  labs(title = paste("Relationship Between Jail Population Rate and Black",
         "Jail Population Rate"),
       x = "Total Jail Population Rate (Unit: 0.01)",
       y = "Black Jail Population Rate (Unit: 0.01)")

```

### Map

The chart is an essential tool for visualizing and understanding the 
distribution of incarceration rates among the White population across different 
counties within Washington State. This map was included to provide a geographic 
perspective on the disparities or uniformities in prison population rates, 
Maps provide a clear visual way to spot regional trends and disparities which 
can be crucial for policy development. Focusing on the White prison population 
allows for the examination of how incarceration rates differ within this 
specific demographic. The map shows varying levels of incarceration rates across 
the state, with some counties exhibiting significantly higher rates than others. 
Counties with darker shades indicate higher average White prison population 
rates, suggesting regions where incarceration is more prevalent within this 
demographic. Conversely, lighter shades indicate lower rates. There are about 3 
southwestern counties have relatively dark green, indicating higher white prison 
population rate. Others are relatively even in white prison population rate.

``````{r}
library(mapproj)
library(maps)

map_wa <- prison_wa %>%
  filter(year < 2017) %>%
  mutate(county = tolower(gsub("\\s.*", "", county_name))) %>%
  filter(!is.na(white_prison_pop_rate)) %>%
  group_by(county) %>%
  summarize(avg_white_rate = mean(white_prison_pop_rate)) %>%
  select(county, avg_white_rate)

county_shape <- map_data("county") %>%
  filter(region == "washington") %>%
  rename(county = subregion) %>%
  left_join(map_wa, by = "county")
  

ggplot(county_shape) +
  geom_polygon(
    mapping = aes(x=long, y=lat, group = group, 
                  fill = avg_white_rate),
    color = "white",
    size = .1
  ) +
  coord_map() +
  scale_fill_continuous(low = "grey90", high = "darkgreen") +
  labs(title = paste("Average White Prison Population Rate in Each County",
                     "in Washington"),
       fill = "Average White Prison Population Rate (Unit: 0.01)", 
       x = "Longitude", 
       y = "Latitude") +
  theme_minimal()
```
